### 2020-03-19 Manuscript/manuscript.tex
small edits and move much of the intro text to conclusion

\documentclass[11pt,letterpaper]{article}
\setcounter{section}{2}
\usepackage{threeparttable}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{pdfpages}
\usepackage{url}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[outercaption]{sidecap}
\doublespacing

@@

Laurentia was a major continent throughout the majority of the Proterozoic. It is hypothesized to have had a central position in both the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. Data from the Slave and Superior provinces of Laurentia provide what is arguably the strongest evidence of differential plate tectonics in the Rhyacian and Orosirian Periods of the Paleoproterozoic Era (2.3 to 1.8 Ga) leading up to the collision of these terranes during the Trans-Hudson orogeny. The collisions of these and other Archean provinces led to the formation of the core of Laurentia. Subsequent crustal growth occurred through multiple intervals of accretionary orogenesis through the late Paleoproterozoic and Mesoproterozoic until the continent-continent collision of the Grenvillian orogeny that was ongoing at the Mesoproterozoic-Neoproterozoic boundary (1.0 Ga). The lead-up to this orogeny was associated with rapid plate motion of Laurentia from high latitudes towards the equator recorded by the Logan Loop and Keweenawan Track of paleomagnetic poles. Following, a return to high-latitudes as constrained by paleomagnetic poles of the Grenville Loop, Laurentia straddled the equator at the time of Cryogenian Snowball Earth glaciation as part of the Rodinia supercontinent. Rifting and passive margin development then isolated Laurentia in the early Paleozoic Era. Subsequent accretionary and collisional orogenesis occurred associated with the Appalachian orogenic cycle with Laurentia first colliding with Avalonia-Baltica to become Laurussia and Laurussia then uniting with Gondwana to form Pangea. While the details of the conjugate continents are better reconstructed for this last Wilson cycle, the broad features of the Trans-Hudson, Grenvillian and Appalachian orogenic cycles bear similarities. In each case, accretionary collision of arc terranes was followed by continent-continent collision. The major difference is that the collisions of the Grenvillian and Appalachian orogenic cycles resulted in relatively minor crustal growth compared to the Trans-Hudson. Break-up following the Grenvillian and Appalachian orogenic cycles occurred along the same margin as collision while the major orogens of the Trans-Hudson orogenic cycle have remained sutured. As a result, Laurentia has been a formidable continent for the past 1.8 billion years whose position and history is key for reconstructing global paleogeography. The evidence from Laurentia provides strong-support for mobile lid plate tectonic processes operating over the past 2.2 billion years.

@@

Laurentia refers to the craton that forms the Precambrian core of North America (Fig. \ref{fig:Laurentia_map}). Laurentia is comprised of multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collision orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.

@@

In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period terminal collision not only resulted in the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.

>>>>

\documentclass[11pt,letterpaper]{article}
\setcounter{section}{3}
\usepackage{threeparttable}
\usepackage{textcomp,marvosym}
\usepackage{amsmath,amssymb}
\usepackage[left]{lineno}
\usepackage{changepage}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{pdfpages}
\usepackage{url}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[outercaption]{sidecap}
\doublespacing

@@

Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provides strong-support for mobile lid plate tectonic processes operating over the past 2.2 billion years.

@@

Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}. As for the Wyoming province, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming province and Medicine Hat blocks were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a,Hrncir2017a}, as well as high-grade tectonism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga lead to rapid amalgamation of the majority of the Laurentia craton.

@@

In the paleogeographic model framework of \cite{Pehrsson2015a}, the collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}). The \cite{Pehrsson2015a} model posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but is also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Siberia, Congo/S\~ao Francisco, West Africa, and Amazonia \citep{Whitmeyer2007a, Pehrsson2015a}.
